---
title: "Distribution of Farmers Markets"
author: "Martin Oberg"
date: "04/03/2021"
output: 
  pdf_document: 
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

This document reports on trends of Farmers Markets accessibility throughout the United States using data collected from https://www.kaggle.com/madeleineferguson/farmers-markets-in-the-united-states.  This is a data set that requires date cleaning and the main aim is project to provide some code for how to do so.  I will then compare market accessibility as measured by number of days that markets are open; having many markets is beneficial to consumers, but if they are only open for a few weeks then it is not reasonable to think that consumers could buy a significant portion of their yearly food at Farmers Markets.

I will also compute the number of markets per capita and compare results.

```{r}
library(tidyverse)
library(lubridate)

data_path = "D:/data/farmers-markets-in-the-united-states/farmers_markets_from_usda.csv"
county_path = "D:/data/farmers-markets-in-the-united-states/wiki_county_info.csv"

data_cols = cols(
  .default = col_character(),
  FMID = col_integer(),
  Season4Date = col_character(),
  Season4Time = col_character(),
  x = col_double(),
  y = col_double()
)

data_full = read_csv(data_path, col_types = data_cols)
cleaned_data = FALSE
data_county = read_csv(county_path)
```


# Data Cleaning

There must have been a lot of manual data entry for this data set because there are a lot of non-standard date entries.  On first glance of the column names one would suspect that Season1-4 corresponds to seasons throughout the year and the various dates and times of market hours.  This might be the case for some markets, however, Season1Date is filled in for most markets and date often extend for more than a year.  I will make the assumption that market dates that extend beyond a year are simply year round markets without intermittent closures as there is no evidence to the contrary. I will also be ignoring columns Season2-4 as the purpose of those columns is unclear.

Another analysis could look at how many markets offer which kinds of products.  That will be left for a different report.

This is a look at the kind of data we have.
```{r pressure, echo=FALSE}
data_full %>% 
  select(FMID, County, State, Season1Date) %>% 
  head(5)
```

The following code fixes problems in data entry.
```{r}
# There are quite a few different formats that dates are recorded in.  Some misspellings, some extra text, some illegal dates (e.g. April 31).
# This will fix them for the date parser.
if (!cleaned_data) {
  data_full[data_full$FMID == 1011963, "Season1Date"] = str_replace(data_full[data_full$FMID == 1011963, "Season1Date"], "Sept.", "September")
  data_full[data_full$FMID == 1008935, "Season1Date"] %>% 
    map_chr(~ str_replace(.x, "Start Date ","")) %>% 
    map_chr(~ str_replace(.x, "End Date ","")) -> data_full[data_full$FMID == 1008935, "Season1Date"]
  data_full[data_full$FMID == 1010153, "Season1Date"] %>% 
    map_chr(~ str_replace(.x, "Octobsr", "October")) -> data_full[data_full$FMID == 1010153, "Season1Date"]
  data_full[data_full$FMID == 1001139, "Season1Date"] %>% 
    map_chr(~ str_replace(.x, "Sept", "September")) -> data_full[data_full$FMID == 1001139, "Season1Date"]
  data_full$Season1Date %>%  
    map_chr(~ str_replace(.x, "09/31", "09/30")) -> data_full$Season1Date
  data_full$Season1Date %>%  
    map_chr(~ str_replace(.x, "11/31", "11/30")) -> data_full$Season1Date
  data_full$Season1Date %>%  
    map_chr(~ str_replace(.x, "04/31", "04/30")) -> data_full$Season1Date
  data_full[data_full$FMID == 1011963, "Season1Date"] = "June 23, 2012 to September 8, 2012"
  data_full[data_full$FMID == 1008935, "Season1Date"] = "01/01/2013 to 12/31/2013"
  data_full[data_full$FMID == 1004852, "Season1Date"] = "June 30, 2011 to October 13, 2011"
}
# There are som string replace functions that are called that should only be run once!
cleaned_data = TRUE
```


And now we can calculate how long each market is open.
```{r}

data_good_dates = data_full %>% 
  select(FMID, State, County, Season1Date) %>% 
  separate(Season1Date, into = c("SeasonBeg", "SeasonEnd"), sep = " to", fill = "right") %>% 
  filter(!is.na(SeasonBeg)) %>% 
  filter(!is.na(SeasonEnd)) %>% 
  filter(nchar(SeasonEnd) > 1 )

data_good_dates %>% 
  mutate(begDate = case_when( !str_detect(SeasonBeg, regex("[:alpha:]")) ~ parse_date_time2(SeasonBeg, "%m/%d/%Y", exact=TRUE), 
                              !str_detect(SeasonBeg, regex("[:digit:]")) ~ parse_date_time(SeasonBeg, orders = c("b", "B"), train=F,drop=T),
                              TRUE ~ parse_date_time(SeasonBeg, orders = c("b %d, %Y", "b %d, %Y", "b %d", "%m/%d/%Y"), exact=F, train=F,drop=T)
                              )) -> markets_by_days
markets_by_days %>% 
  mutate(endDate = case_when( !str_detect(SeasonEnd, regex("[:alpha:]")) ~ parse_date_time2(SeasonEnd, "%m/%d/%Y", exact=TRUE), 
                              !str_detect(SeasonEnd, regex("[:digit:]")) ~ parse_date_time(SeasonEnd, orders = c("b", "B"), train=F,drop=T),
                              TRUE ~ parse_date_time(SeasonEnd, orders = c("b %d, %Y", "b %d, %Y", "b %d", "%m/%d/%Y"), exact=F, train=F,drop=T)
                              )) -> markets_by_days
  
# Fix missing years in begDate
markets_by_days %>% 
  mutate(begDate = if_else(year(begDate) == 0, update(begDate, year = year(endDate)), begDate)) %>% 
  mutate(DaysOpen = map2_int(begDate, endDate, ~ as.integer(.y-.x))) -> markets_by_days 
# Fix missing years in endDate
markets_by_days %>% 
  mutate(endDate = if_else(year(endDate) == 0, update(endDate, year = year(begDate)), endDate)) %>% 
  mutate(DaysOpen = map2_int(begDate, endDate, ~ as.integer(.y-.x))) -> markets_by_days 
  #filter(abs(nDays) > 366)

# Some markets have mdy for SeasonBeg, but only month for SeasonEnd.  This 
markets_by_days %>% 
  #filter(nDays < 0)
  mutate(endDate = if_else(month(begDate) > month(endDate), 
                           update(endDate, year = year(begDate) + 1),
                           endDate)) %>% 
  mutate(DaysOpen = map2_int(begDate, endDate, ~ as.integer(.y-.x))) -> markets_by_days

# only outliers now are DoysOpen > 365.
markets_by_days %>% 
  filter(DaysOpen <0) 

#markets open for more than a year will be considered to be open for a full year
markets_by_days %>% 
  mutate(DaysOpen = if_else(DaysOpen > 365, as.integer(365), DaysOpen)) -> markets_by_days
  
head(markets_by_days)
```


```{r}
markets_by_days = markets_by_days %>% 
  select(FMID, State, County, DaysOpen)

markets_by_days %>% 
  ggplot(aes(x=DaysOpen)) +
  geom_histogram()
```

```{r}
markets_by_days %>% nrow()
```
```{r}
markets_by_days %>% filter(DaysOpen >= 350) %>% nrow() / 5563
```
```{r}
markets_by_days %>% filter(DaysOpen < 350) %>% select(DaysOpen) %>% summary()
```

Roughly 13% of markets are open year-round, or nearly year round.  The rest are normally distributed around an average of 148 days, about 5 months.


```{r}
library(tmap)
library(spData)
library(spDataLarge)
data(us_states)

state_population = 
  data_county %>% 
  group_by(State) %>% 
  summarise(pop = sum(population))

norm_days = markets_by_days %>% 
  group_by(State) %>% 
  summarise(TotalDays = sum(DaysOpen %% 366, na.rm=T)) %>% 
  right_join(state_population) %>% 
  mutate(normDays = TotalDays/pop*1000) 

norm_days_states = 
  us_states %>% 
  right_join(norm_days, by=c("NAME" = "State"))
```


```{r}
tm_shape(norm_days_states, projection = 2163) + 
  tm_polygons("normDays" , title = "Market Days per 1000") + 
  tm_layout(frame = FALSE) +
  tm_layout(main.title = "Number of Market days in the US", title.size = 1.5,main.title.position="center")
```

With categories breaks of 0, 5, 10, and 15 market days open per 1,000 population, the state of Farmers Markets across the country looks rather grim.


```{r}
tm_shape(norm_days_states, projection = 2163) + 
  tm_polygons("normDays" , title = "Market Days per 1000", breaks=c(0,3,6,11,15)) + 
  tm_layout(frame = FALSE) +
  tm_layout(main.title = "Number of Market days in the US", title.size = 1.5,main.title.position="center")
```

Looking at the number of days that Farmers Markets are open shows how often throughout the year consumers are able to shop at markets.  Ignoring population means that markets may be busy or perhaps in a neighboring county.  I was surprised to find an almost inverse effect of latitude.


## Number of Markets by State

If we simply look at the number of markets in each state we can avoid all of the date cleaning that was required in the previous section.

```{r}
data_full %>% 
  group_by(State) %>% 
  summarise(N = n()) %>%
  right_join(state_population) %>% 
  mutate(NperCapita = N/pop*100000) -> markets_by_states

```

```{r}
state_market_counts = 
  us_states %>% 
  right_join(markets_by_states, by=c("NAME" = "State"))

tm_shape(state_market_counts, projection = 2163) + 
  tm_polygons("N" , title="Number of Markets") + 
  tm_layout(frame = FALSE) +
  tm_layout(main.title = "Number of Farmers Markets in the US", title.size = 1.5,main.title.position="center")

```

```{r}

tm_shape(state_market_counts, projection = 2163) + 
  tm_polygons("NperCapita", title="Markets per 10,000") + 
  tm_layout(frame = FALSE) +
  tm_layout(main.title = "Number of Farmers Markets in the US per Capita", title.size = 1.5,main.title.position="center")
```

Looking at the total number of markets in each state show that New York and California clearly have the most, however, per capita they are well below average.  Looking at these states at the county level could be useful.  Vermont stands out in the leader of markets per capita and in number of open market days.  Oregon, Montana, Wyoming, New Hampshire, and Maine are leaders in open market days and also stand out in number of markets per capita.  
